### 2009-05-29 response-to-reviews2.txt

We do hope that the sophistication of the constraint schemes used with
these "tree transducer" models will increase over time. Eventually we
hope to develop alignment constraints, simultaneous alignment-and-fold
constraints, and other constraints (and/or sampling schemes) to enable
efficient analysis of multiple phylogenetically-related sequences. The
present work represents a first attempt at using these sorts of models
for more than two sequences, but we hope it will not be the last
attempt, and we acknowledge room for improvement.

> It is still not clear eenough what the authors' model is, though it is 
> supposed to be a main result. Page 7 states "a prescriptive formalism 
> for transducers is given in Text S1". Relegating the deliverables of a 
> manuscript to supplementary material is not an appropriate use of 
> supplementary material; a reader should not have to refer to a second 
> manuscript to understand the main points. The authors need to find a 
> way to explain their main results in the actual manuscript. There is 
> still much vagueness and redundancy in this draft, which could be 
> removed and replaced with more relevant explanation. For example, in 
> this same section where results are relegated to supplementary 
> material, there's a whole redundant paragraph (p.8) on how conditional 
> models represent directed edges; but this concept has already been 
> explained sufficiently. This is followed (bottom p. 8, "Consider a 
> phylogenetic tree...") by yet another statement of the same concept of 
> using conditional models, still in
> terms too vague and conceptual to enable reproduction of the work.

Thank you for pointing this out.  We have removed the reference to the
supplementary material in this section and have attempted to make the
main text a self-contained description of the composition algorithm,
including its key corollaries (elimination of null states, and DP
inference).  The presentation of the composition algorithm in the
supplement now merely reprises the presentation in the main text using
more formal notation. We hope this is an appropriate use of the
supplementary information, but we welcome additional suggestions to
improve the flow of the manuscript for the reader.

We have attempted to remove redundant text, including the two
paragraphs which the reviewer notes, and other sentences.

@@

Thank you for the feedback; we have attempted to revise the paper
accordingly such that it is completely self-contained and
reproducible.  We have given complete specifications of the CYK,
Inside, Outside, and CYK traceback algorithms for three sequences,
constrained by a pre-determined set of folds for each sequence ("fold
envelopes").  They are written in pseudocode which corresponds closely
to our actual implementation in the program (INDIEGRAM) which we
describe.

We have entirely replaced the "loopy DP" algorithm with an algorithm
for exactly eliminating null cycles from general SCFGs.  Per the
reviewer's recommendation, the full description of this algorithm is
described in the main text ("Exact elimination of null cycles in
SCFGs") of the revised manuscript.  This new algorithm avoids the
increased time complexity associated with the iterative inference of
"loopy DP."

@@

We agree that these reconstructions were not very informative.
Unfortunately these were the only sequences that we could find which
were short enough to be tractable on the hardware we had available. We
now report simulation results instead, as suggested.

We thank the reviewer for pointing out errors in the original
figures. These were due to mistakes made during manual conversion of
indiegram's output parse trees into Stockholm-format alignments and a
bug in the code used for maximum-likelihood reconstruction of
sequences (the xrate program). We have now automated this process and
fixed the bug, verifying that this has eliminated the errors
(in fact this verification is incorporated into our new benchmark).

We removed the tRNA and nanos reconstructions and posted (corrected)
versions online at the following URL (which is linked from the
indiegram webpage),

http://biowiki.org/IndiegramExamples

although we note in the text that the reconstructions were too easy
because of short branch lengths.

@@

> p.6 bottom: pair HMMs are not really "standard"; they are not widely 
> used or known. (There is still too much of a general tendency in the 
> manuscript to use technically narrow jargon in place of actual 
> explanation, and assume the reader will be able to fill in the gaps. 
> Sometimes, though, the revision gets it right: for example, other than 
> this "standard pair HMM" reference, the authors do an excellent job on 
> pp6-7 introducing the concept and motivation of conditional 
> probability models first, and *then* connecting to the linguistics 
> literature and the jargon term "parse-tree transducer".)

Thank you for pointing out this oversight.  We briefly describe what a
Pair HMM is in the revised manuscript.

> p.7 why is the model that generates P(X) at the root a "singlet 
> transducer"? It's not transducing anything (it's not conditionally 
> dependent on an input parse tree).

By "singlet transducer" we mean the same thing as is often referred to
as a "single-sequence SCFG".  Instead of "a model that generates a
parse tree X", we use the contrivance of saying "a model that inputs a
(dummy) empty parse tree, and outputs a parse tree X".  It's awkward,
but it keeps the notation straight: it allows us to be consistent
throughout the manuscript. Rather than always having to specify
whether a model is generative (SCFG) or transductive (input-output),
we can treat all models as transducers, if we just think of generative
models as transducers with a (dummy) input that is always zero/empty.

@@

> p.13 The DP inference algorithm section says "the standard CYK ... 
> algorithms fail". But then it says "Maximum likelihood inference is 
> performed with a CYK algorithm", and even "maximum likelihood 
> inference can therefore be performed with a standard three sequence 
> CYK algorithm". Which is it?

Thank you for pointing out this inconsistency, which we have clarified
in the revised text.  Using the standard CYK algorithm on the composed
model fails; however, the model can be simply modified as described in
the revised text to permit standard CYK to work properly.

> p.14 It seems possible that the problem with "null cycles" (residues 
> in ancestors that leave no homologous residues in observed sequences) 
> may simply go away if you view the problem as a missing data problem, 
> and that the likelihood being inferred is the likelihood of alignment 
> columns needed to account for the observed data (marginalized over 
> missing columns needed to account for null cycles) rather than the 
> joint likelihood of the "observed" columns and the "null cycle" 
> columns. This is what dropped out of the equations in the phylogenetic 
> inference described by Rivas and Eddy (2008) PLoS CB 4:e1000172, for 
> example.

We completely agree. Such an algorithm was initially a goal of this
work, but was abandoned. We thank the reviewer for challenging us to
attack it afresh; a verbal presentation of the algorithm we came up
with is given in the revised text, with a full formal description
(including notation) in the supplement. (Since this is a standalone
result that can be applied to any SCFG, we think it fits better as an
appendix in the supplement; the obvious cost is omitting some detail
from the main paper, but the PLoS Authors' Guidelines and the
reviewer's preference for including more detail in the manuscript are
potentially in conflict on the best approach to take here. We welcome
suggestions for improving the flow generally.)

As a technical aside, the primary difficulty with SCFG null cycle
elimination is that while this marginalization is reasonably
straightforward for independent-sites models and HMMs, the possible
presence of null bifurcations complicates matters for SCFGs.  Indeed,
no *closed form* elimination is possible (due to bifurcations, such an
elimination would involve solving simultaneous nonlinear equations, as
explained in the supplement) but an iterative approximation
(e.g. Newton-Raphson) appears adequate.

### 2009-05-29 paper.tex
Revised supplement; Ian's changes to Rob's changes

% Title must be 150 words or less
\begin{flushleft}
  {\Large
    \textbf{Evolutionary triplet models of structured RNA}
  }
\\
Robert K. Bradley$^{1}$, 
Ian Holmes$^{1,2,\ast}$
\\
\textbf{1} Biophysics Graduate Group, University of California, Berkeley, CA, USA
\\
\textbf{2} Department of Bioengineering, University of California, Berkeley, CA, USA
\\
$\ast$ E-mail: indiegram@postbox.biowiki.org
\end{flushleft}

@@

We describe a ``transducer composition'' algorithm for extending
pairwise statistical models of RNA structural evolution to models of
multiple sequences related by a phylogenetic tree.  This algorithm
draws on formal models of computational linguistics as well as the
1985 ``protosequence'' algorithm of David Sankoff.  The output of the
composition algorithm is a multiple-sequence stochastic context-free
grammar.  We describe dynamic programming algorithms, which are robust
to null cycles and empty bifurcations, for parsing this grammar.
Example applications include structural alignment of non-coding RNAs,
propagation of structural information from an
experimentally-characterized sequence to its homologs, and inference
of the ancestral structure of a set of diverged RNAs.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\newpage
\section*{Author Summary}
% to do: write this

@@

% SCFGs for RNA
An inspection of RNA alignments, such as those in the \RFAM\ database
\cite{GriffithsJonesEtAl2003}, suggests that an evolutionary model for
RNA structure must eventually include multiple layers of detail: point
substitutions, covariant substitutions of base-pairs
\cite{HancockTautzDover88,LeontisEtAl2002}, indels
\cite{TakeshiTsutomu2008}, local changes in secondary structure such
as helix slippage \cite{HancockDover90}, and changes in domain
structure \cite{SmithEtAl2008}.  Stochastic context-free grammars
(SCFGs), which can efficiently model the long-range correlations
generated by RNA secondary structures, are natural probabilistic models of such
phenomena and have been used for ncRNA homology detection
\cite{Eddy94,Durbin98,KleinEddy2003,NawrockiEddy2007}, gene prediction
\cite{RivasEddy99,PedersenEtAl2006}, folding
\cite{KnudsenHein2003,DowellEddy2004} and alignment
\cite{Holmes2005,DowellEddy2006,BradleyPachterHolmes2008}.

% Two parts: pairwise & multiple
By analogy with substitution processes, which are well-understood \cite{Felsenstein2003},
we may split the ancestral reconstruction problem into two halves.
The first half is the development of a {\bf pairwise model},
describing the probability distribution $P(Y|X)$ of a descendant ($Y$) conditional on its immediate ancestor ($X$).
In substitution processes, the pairwise model is a conditional substitution matrix.
Often (but not always) the pairwise model is derived from an instantaneous model of change,
i.e., a continuous-time Markov chain (parametrized by a rate matrix);
obtaining the transition probabilities of this chain (via exponentiation of the rate matrix)
yields a pairwise model whose parameters are smoothly-varying functions of a finite time parameter $T$.
A pairwise model represents an individual branch of a phylogenetic tree, with $T$ representing the length of that branch.

% (second part: multiple)
The second half of the problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
In a typical situation, the sequences at the leaves of the tree are observed, but those at internal nodes are unobserved.
Questions of interest then include:
\begin{description}
\item[A.] What is the likelihood for the observed sequence data?
\item[B.] Can we sample (find the mode, take moments, etc.) from the posterior distribution of the unobserved sequence at the root node?
\item[C.] Can we sample (etc.) from the posterior of the unobserved sequences at the other internal nodes?
\item[D.] Can we estimate summaries of the evolutionary history, such as
  the number of substitution events on each branch (for a substitution
  model), the alignment (for a model which includes indels), or
  changes in the underlying structure (for a model of RNA structure)?
\end{description}

For substitution models, there has been extensive work focused on answering each of these questions.
Given a pairwise substitution model, questions A and B can be
answered exactly by Felsenstein's pruning algorithm
\cite{Felsenstein81} and question C can be answered by the peeling
algorithm (first presented for pedigree analysis by Elston and Stewart \cite{ElstonStewart71}).
The estimation of evolutionary histories (question D) has been
addressed by exact summarization \cite{HolmesRubin2002b} and sampling
\cite{Nielsen2001} approaches.
Many of these questions and answers can be unified by an
appropriate framework;
for example, another representation of answers A-C
is that the ``pruning'' and ``peeling'' algorithms (combined) are just
the sum-product algorithm on a directed graphical model \cite{Pearl82},
yielding exact marginal distributions for unobserved variables.
Graphical models also suggest general-purpose sampling approaches
in addition to the exact sum-product algorithm.

% We want a multiple SCFG. Not a pairwise SCFG...
The two halves of the reconstruction problem --- developing a pairwise model and then extending it to multiple sequences --- are largely independent.
Felsenstein's pruning algorithm, for example, is essentially blind to the parametric form of the pairwise substitution model;
it just assumes that a substitution model is provided for every branch.
Subsequent models developed by other researchers can be plugged into the pruning algorithm without modification \cite{Yang94b,WhelanGoldman2001}.

@@

A stochastic model for the evolution of one sequence (the ancestor, $X$) into
another (the descendant, $Y$) over an interval of time ($T$) can be described by a
joint distribution, $P (X, Y | T)$.
This joint distribution can be factored as $P (X) \cdot P (Y | X, T)$, 
where $P(X)$ is the marginal distribution over ancestral sequences
and $P(Y | X, T)$ is the conditional distribution over descendant sequences given an ancestral sequence.
In terms of phylogenetics, the conditional distribution $P(Y | X, T)$ describes the evolution $X \stackrel{T}{\to} Y$
along a branch of length $T$.

@@

A similar formalism can be used to describe the evolution of whole
sequences with indels.
Suppose that the joint distribution $P (X, Y | T)$ is the distribution
modeled by a pair hidden Markov model (Pair HMM) \cite{Durbin98},
a probabilistic model of the evolution of two sequences under the
approximation that only adjacent characters are directly correlated,
and the marginal $P(X)$ is the distribution of a single-sequence HMM,
a probabilistic model of single sequences under the same approximation.
The conditional distribution $P(Y | X, T)$ then corresponds to a conditional Pair HMM,
a finite-state machine which transforms one sequence (the input, $X$) into another (the output, $Y$).
Following computational linguists, we call this conditionally-normalized state machine a
 {\bf string transducer} or simply a {\bf transducer}
\cite{BradleyHolmes2007}.
Because of its conditional normalization, this state machine is
distinct from a standard Pair HMM.
A Pair HMM has two outputs $X$ and $Y$ and emits symbols to both of those outputs,
while a transducer absorbs symbols from the input $X$ and emits
symbols to the output $Y$.
Despite this distinction, Pair HMMs and transducers share very similar inference algorithms;
for example, $P(Y | X, T)$ is computed using a direct analogue of the Forward algorithm \cite{Durbin98}.

@@

Transducers can have states of type $\Sstart$, $\Send$,
$\Swait$, $\Sinsert$ and $\Smatch$.
The first three state types, $\Sstart$, $\Send$ and $\Swait$, are null:
they do not emit or absorb any symbols and are required solely for
``organizational'' purposes (see following section).
Two types of states can emit and/or absorb symbols, $\Sinsert$ and $\Smatch$.
An $\Sinsert$ state emits a symbol to the output without absorbing anything.
A $\Smatch$ state absorbs a symbol on the input and either emits
the same symbol to the output, substitutes a different output symbol,
or emits no output symbol at all, the last corresponding to a deletion.

As stated above, the Pair SCFG must be ``conditionally normalized''
so that models can be chained together,
extending the pairwise model to multiple sequences.
The transformation rules are partitioned into co-normalized groups;
within each group, the rule probabilities must sum to one.
In a jointly-normalized Pair SCFG, each group corresponds to the set of all rules that can be applied to a given nonterminal
(i.e., all outgoing transitions from a particular state).
In a conditionally-normalized Pair SCFG, in contrast,
each co-normalized group includes all rules that can be applied to a
given nonterminal {\em for a given set of absorbed symbols}.
%(In practice it is often expedient to relax these normalization guidelines somewhat,
%while preserving the overall probabilistic normalization over output parse trees.
%For example, the normalization of states $B$ and $B_p$ in \tabref{tkfstbranch} occurs on transitions into, rather than out of, those states.)

@@

We therefore need an algorithm to efficiently construct the ``state graph'' of the
ensemble model, consisting of a list of accessible states and the
possible transitions between them. 
By analogy with algorithms for uninformed graph search in artificial
intelligence, the transition graph of the ensemble can be constructed
by an uninformed depth-first search, where at each step of the search
we obtain the next possible ensemble states by changing the state of one or
more of the singlet or branch transducers. Beginning with the entire
ensemble in state $\Sstart$, the depth-first search of states
continues until all nodes are in state $\Send$.

The allowed transitions of the ensemble can be categorized as follows:
\begin{enumerate}
\item {\bf Null Transition}:
  A branch transducer makes a transition into a $\Swait$ state, with no terminal emission or bifurcation.
\item {\bf Terminal Emission}:
  A singlet or branch transducer makes a transition into a state of type $\Sinsert$,
  emitting left and/or right terminal symbols (e.g., a single base or base-pair).
  These symbols are absorbed by the immediately-descended transducers,
  which are pushed into states of type $\Smatch$
  and may themselves emit terminal symbols
  that will be absorbed by {\em their} descendant transducers.
  This continues down the tree: The terminal symbols are passed from parents to children to grandchildren
  (albeit possibly being replaced by other terminal symbols as they
  are propagated down)
  and they propel branch transducers into $\Smatch$ states as they go.
  Eventually, the cascade of emitted terminal symbols stops when all the symbols have been deleted
  or when the cascade reaches the leaves of the tree.
\item {\bf Bifurcation}:
  A singlet or branch transducer makes a transition into a state of type $\Sinsert$ that spawns
  left and/or right nonterminal states.
  These nonterminals are processed recursively down the tree, just as in a
  terminal emission (conceptually, a bifurcation is a ``nonterminal emission'').
  As with terminal emissions, absorption of nonterminal emissions propels
  descendant transducers into $\Smatch$ states,
  making transitions which may themselves propagate nonterminals
  further down the tree.
  A biologically-relevant example of a bifurcation is the insertion of a stem
  into an ancestral RNA structure,
  which may then be conserved or deleted in the descendant structures.
\item {\bf End Transition}:
  The singlet transducer at the root makes a transition to the $\Send$ state,
  pushing all the descendant branch transducers into $\Send$ states and
  terminating the current branch of the parse tree.
\end{enumerate}

Co-ordination between the various branch machines is achieved by
specifying an ordering on the nodes and by having branch transducers ``pause'' in $\Swait$
states while waiting to absorb a symbol from the node above.  
Only one transducer is allowed to make a spontaneous transition at a
time. If this transition corresponds to a terminal
emission or a bifurcation, then this may force descendant transducers
into making reactive transitions.

The four types of allowed transitions listed above can be formalized
as follows.
Let the total order on the nodes correspond to a preorder traversal of
the tree, from which it follows that $m \prec n$ if $m$ is ancestral to $n$.
Let ${\cal T}_m$ denote the singlet or branch transducer which emits symbols to node $m$.
Transducer ${\cal T}_m$ changes state if and only if one of the
following three mutually-exclusive conditions holds:
\begin{description}
\item{Type 1:} Transducer ${\cal T}_m$ is not in a $\Swait$ state,
  while all its successor transducers ${\cal T}_n$ are in $\Swait$
  states (where $m \prec n$).  ${\cal T}_m$ is free to make any transition.
\item{Type 2:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer emits a symbol, forcing ${\cal T}_m$ into a $\Smatch$ state so it can absorb that symbol.
\item{Type 3:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters the $\Send$ state, forcing ${\cal T}_m$ into the $\Send$ state as well.
\end{description}
These possible transitions of the ensemble generate multiple alignments as follows:
\begin{enumerate}
\item
  The singlet transducer and all branch transducers begin in their respective $\Sstart$ states.
\item \label{item:InitialWindback}
  Before any residues can appear at the root, the branch transducers all ``wind back'' into $\Swait$ states,
  via type-1 transitions. This occurs in reverse order (i.e.,a postorder traversal of the tree).
\item During this initial windback, clade-specific insertions can occur.
  This process is described in detail at step \ref{item:CladeInsertion}.
\item \label{item:SingletStart}
  With all the branch transducers wound back into $\Swait$ states,
  the singlet transducer makes a (type-1) transition into an $\Sinsert$ state, emitting a symbol to the sequence at the root node.
\item The transducers on outgoing branches from the root then make (type-2) transitions into $\Smatch$ states,
  either copying the root symbol to their own outputs,
  substituting it for a different symbol or staying silent (this silence corresponds to a clade-specific deletion;
  in our formalism, both substitutions and deletions are handled by $\Smatch$ states.)
\item The transducers on branches one step away from the root then process the symbols which reached them
  (if any did), followed by transducers on branches two steps away from the root, then three steps, and so on
  (these can all be regarded as occurring simultaneously, in a single ``wave'' or ``cascade'' of emissions).
\item \label{item:Windback}
  Eventually the emitted symbols are propagated, via type-2 transitions,
  all the way to the tips of the tree (if they survived)
  or to the nodes where they were deleted (if they did not survive).
  The wave of type-2 transitions has left a lot of branch transducers in $\Sinsert$ and $\Smatch$ states.
\item The branch transducers then, in postorder, each wind back into $\Swait$ states, just as at step \ref{item:InitialWindback}.
  (These windback transitions can be collapsed into a single ensemble transition, as with the emission cascade;
  however, the windback may be interrupted by clade-specific insertions; see below.)
\item \label{item:CladeInsertion}
  During the postorder windback, each branch transducer gets an opportunity to generate a new symbol (via type-1 transitions to $\Sinsert$ states).
  (If such a transition to $\Sinsert$ occurs, it corresponds to a clade-specific insertion.
  This insertion is propagated down the tree via a wave of type-2 transitions, as above,
  then we go back to step \ref{item:Windback}.)
\item Eventually, the entire ensemble has wound back, so that every transducer is in a $\Swait$ state except the singlet transducer at the root,
  which is still in an $\Sinsert$ state.
  At this point, all clade-specific insertions have been processed.
\item The singlet transducer now makes another type-1 transition.

>>>>

% Title must be 150 words or less
\begin{flushleft}
  {\Large
    \textbf{\titlestring}
  }
\\
\authorstring
\end{flushleft}

@@

We describe a ``transducer composition'' algorithm for extending
pairwise statistical models of RNA structural evolution to models of
multiple sequences related by a phylogenetic tree.  This algorithm
draws on formal models of computational linguistics as well as the
1985 protosequence algorithm of David Sankoff.  The output of the
composition algorithm is a multiple-sequence stochastic context-free
grammar.  We describe dynamic programming algorithms, which are robust
to null cycles and empty bifurcations, for parsing this grammar.
Example applications include structural alignment of non-coding RNAs,
propagation of structural information from an
experimentally-characterized sequence to its homologs, and inference
of the ancestral structure of a set of diverged RNAs.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\newpage
\section*{Author Summary}

A number of leading methods for bioinformatics analysis of noncoding RNA
 use probabilistic grammars as models for pairwise alignments.
We show that any such pairwise grammar can be extended to an entire phylogeny,
 by considering the pairwise grammar as a machine (a ``transducer'') that locally transforms one RNA structure into another
 (thereby modeling a single parent-child relationship in the tree).
In addition to current applications such as genefinding, homology detection, alignment and secondary structure prediction,
 this should permit statistical reconstruction of RNA genes ancestral to many present-day sequences.
We describe algorithms, software implementations, and an elementary benchmark wherein
 the complexity and accuracy of three-way RNA reconstruction is measured as the length of one branch is varied.
In the Discussion we consider how the three-way RNA alignment-reconstruction-folding algorithm, which is currently very expensive computationally,
 might be made more efficient so that larger phylogenies could be considered.

@@

% SCFGs for RNA
An inspection of RNA alignments, such as those in the \RFAM\ database
\cite{GriffithsJonesEtAl2003}, suggests that an evolutionary model for
RNA structure must eventually include multiple layers of detail: point
substitutions, covariant substitutions of base-pairs
\cite{HancockTautzDover88,LeontisEtAl2002}, indels
\cite{TakeshiTsutomu2008}, local changes in secondary structure such
as helix slippage \cite{HancockDover90}, and changes in domain
structure \cite{SmithEtAl2008}.  Stochastic context-free grammars
(SCFGs), which can efficiently detect the long-range correlations
of RNA base-pairing structures, are natural probabilistic models of such
phenomena and have been used for ncRNA homology detection
\cite{Eddy94,Durbin98,KleinEddy2003,NawrockiEddy2007}, gene prediction
\cite{RivasEddy99,PedersenEtAl2006}, folding
\cite{KnudsenHein2003,DowellEddy2004} and alignment
\cite{Holmes2005,DowellEddy2006,BradleyPachterHolmes2008}.

% Two parts: pairwise & multiple
By analogy with models of substitution processes, which are well-understood \cite{Felsenstein2003},
we may take the problem of building phylogenetic models of RNA evolution and split it into two halves.
The first half is the development of a {\bf pairwise model},
describing the probability distribution $P(Y|X)$ of a descendant ($Y$) conditional on its immediate ancestor ($X$).
In substitution processes, the pairwise model is a conditional substitution matrix.
Often (but not always) the pairwise model, representing a finite evolutionary time $T$,
is derived from an instantaneous model of change over an infinitesimal time interval,
i.e., a continuous-time Markov chain (parametrized by a rate matrix).
Obtaining the transition probabilities of this chain (via exponentiation of the rate matrix)
yields a pairwise model whose parameters are smoothly-varying functions of $T$.
A pairwise model represents an individual branch of a phylogenetic tree, with $T$ representing the length of that branch.

% (second part: multiple)
The second half of the phylogenetic modeling problem involves extending the model (and related inference algorithms) from a single branch to a complete phylogeny,
i.e., from a pairwise model of two sequences to a {\bf multiple-sequence model} of many sequences.
In a typical situation, the sequences at the leaves of the tree are observed, but those at internal nodes are unobserved.
Questions of interest then include:
\begin{description}
\item[A.] What is the likelihood for the observed sequence data?
\item[B.] Can we sample (find the mode, take moments, etc.) from the posterior distribution of the unobserved sequence at the root node?
\item[C.] Can we sample (etc.) from the posterior of the unobserved sequences at the other internal nodes?
\item[D.] Can we estimate summaries of the evolutionary history, such as
  the number of substitution events on each branch (for a substitution
  model), the alignment (for a model which includes indels), or
  changes in the underlying structure (for a model of RNA structure)?
\end{description}

For substitution models, there has been extensive work focused on answering each of these questions.
Given a pairwise substitution model, questions A and B can be
answered exactly by Felsenstein's pruning algorithm
\cite{Felsenstein81} and question C can be answered by the peeling
algorithm (first presented for pedigree analysis by Elston and Stewart \cite{ElstonStewart71}).
The estimation of evolutionary histories (question D) has been
addressed by exact summarization \cite{HolmesRubin2002b} and sampling
\cite{Nielsen2001} approaches.
Another representation of answers A-C
is that the pruning and peeling algorithms (combined) are just
the sum-product algorithm on a directed graphical model \cite{Pearl82},
yielding exact marginal distributions for unobserved variables.
Graphical models also suggest general-purpose sampling approaches
in addition to the exact sum-product algorithm.

% We want a multiple SCFG. Not a pairwise SCFG...
The two halves of the reconstruction problem --- developing a pairwise model and then extending it to multiple sequences --- are largely independent.
Felsenstein's pruning algorithm, for example, is essentially blind to the parametric form of the pairwise substitution model;
it just assumes that a substitution matrix is provided for every branch.
Subsequent models developed by other researchers can be plugged into the pruning algorithm without modification \cite{Yang94b,WhelanGoldman2001}.

@@

A stochastic model for the evolution of one sequence (the ancestor, $X$) into
another (the descendant, $Y$) over an interval of time ($T$) can be described by a
joint distribution, $P (X, Y | T)$.
This joint distribution can be factored, $P (X) \cdot P (Y | X, T)$, 
where $P(X)$ is the marginal distribution over ancestral sequences
and $P(Y | X, T)$ is the conditional distribution over descendant sequences given an ancestral sequence.
In terms of phylogenetics, the conditional distribution $P(Y | X, T)$ describes the evolution $X \stackrel{T}{\to} Y$
along a branch of length $T$.

@@

A similar formalism can be used to describe the evolution of whole
sequences with indels.
Suppose that the joint distribution $P (X, Y | T)$ is the distribution
modeled by a pair hidden Markov model (Pair HMM) \cite{Durbin98},
a probabilistic model of the evolution of two sequences under the
approximation that only adjacent characters are directly correlated,
and the marginal $P(X)$ is the distribution of a single-sequence HMM,
a probabilistic model of single sequences under the same approximation.
The conditional distribution $P(Y | X, T)$ then corresponds to a conditional Pair HMM,
a discrete-state machine which transforms one sequence (the input, $X$) into another (the output, $Y$).
Following computational linguists, we call this conditionally-normalized state machine a
 {\bf string transducer} or simply a {\bf transducer}
\cite{BradleyHolmes2007}.
Because of its conditional normalization, this state machine is
distinct from a standard Pair HMM.
A Pair HMM has two outputs $X$ and $Y$ and emits symbols to both of those outputs,
while a transducer absorbs symbols from the input $X$ and emits
symbols to the output $Y$.
Despite this distinction, Pair HMMs and transducers share very similar inference algorithms;
for example, $P(Y | X, T)$ is computed using a direct analogue of the Forward algorithm \cite{Durbin98}.

@@

Transducers can have states of type $\Sstart$, $\Send$,
$\Swait$, $\Sinsert$ and $\Smatch$.
The first three state types, $\Sstart$, $\Send$ and $\Swait$, are null:
they do not emit or absorb any symbols and are required solely for
organizational purposes (see following section).
Two types of states can emit and/or absorb symbols, $\Sinsert$ and $\Smatch$.
An $\Sinsert$ state emits a symbol to the output without absorbing anything.
A $\Smatch$ state absorbs a symbol on the input and either emits
the same symbol to the output, substitutes a different output symbol,
or emits no output symbol at all, the last corresponding to a deletion.

As stated above, the Pair SCFG must be conditionally normalized
so that models can be chained together,
extending the pairwise model to multiple sequences.
The transformation rules are partitioned into co-normalized groups;
within each group, the rule probabilities must sum to one.
In a jointly-normalized Pair SCFG, each group corresponds to the set of all rules that can be applied to a given nonterminal
(i.e., all outgoing transitions from a particular state).
In a conditionally-normalized Pair SCFG, in contrast,
each co-normalized group includes all rules that can be applied to a
given nonterminal {\em for a given set of absorbed symbols}.
%(In practice it is often expedient to relax these normalization guidelines somewhat,
%while preserving the overall probabilistic normalization over output parse trees.
%For example, the normalization of states $B$ and $B_p$ in \tabref{tkfstbranch} occurs on transitions into, rather than out of, those states.)

@@

We therefore need an algorithm to efficiently construct the state graph of the
ensemble model, consisting of a list of accessible states and the
possible transitions between them. 
By analogy with algorithms for uninformed graph search in artificial
intelligence, the transition graph of the ensemble can be constructed
by an uninformed depth-first search, where at each step of the search
we obtain the next possible ensemble states by changing the state of one or
more of the singlet or branch transducers. Beginning with the entire
ensemble in state $\Sstart$, the depth-first search of states
continues until all nodes are in state $\Send$.

The allowed transitions of the ensemble can be categorized as follows:
\begin{description}
\item[{\bf Null Transition}:]
  A branch transducer makes a transition into a $\Swait$ state, with no terminal emission or bifurcation.
\item[{\bf Terminal Emission}:]
  A singlet or branch transducer makes a transition into a state of type $\Sinsert$,
  emitting left and/or right terminal symbols (e.g., a single base or base-pair).
  These symbols are absorbed by the immediately-descended transducers,
  which are pushed into states of type $\Smatch$
  and may themselves emit terminal symbols
  that will be absorbed by {\em their} descendant transducers.
  This continues down the tree: The terminal symbols are passed from parents to children to grandchildren
  (albeit possibly being replaced by other terminal symbols as they
  are propagated down)
  and they propel branch transducers into $\Smatch$ states as they go.
  Eventually, the cascade of emitted terminal symbols stops when all the symbols have been deleted
  or when the cascade reaches the leaves of the tree.
\item[{\bf Bifurcation}:]
  A singlet or branch transducer makes a transition into a state of type $\Sinsert$ that spawns
  left and/or right nonterminal states.
  These nonterminals are processed recursively down the tree, just as in a
  terminal emission (conceptually, a bifurcation is a ``nonterminal emission'').
  As with terminal emissions, absorption of nonterminal emissions propels
  descendant transducers into $\Smatch$ states,
  making transitions which may themselves propagate nonterminals
  further down the tree.
  A biologically-relevant example of a bifurcation is the insertion of a stem
  into an ancestral RNA structure,
  which may then be conserved or deleted in the descendant structures.
\item[{\bf End Transition}:]
  The singlet transducer at the root makes a transition to the $\Send$ state,
  pushing all the descendant branch transducers into $\Send$ states and
  terminating the current branch of the parse tree.
\end{description}

Co-ordination between the various branch machines is achieved by
specifying an ordering on the nodes and by having branch transducers pause in $\Swait$
states while waiting to absorb a symbol from the node above.  
Only one transducer is allowed to make a spontaneous transition at a
time. If this transition corresponds to a terminal
emission or a bifurcation, then this may force descendant transducers
into making reactive transitions.

The four types of allowed transitions listed above can be formalized
as follows.
Let the total order on the nodes correspond to any preorder traversal of
the tree; thus, ``$m$ is ancestral to $n$'' is sufficient-but-not-necessary for ``$m \prec n$''.
Let ${\cal T}_m$ denote the singlet or branch transducer which emits symbols to node $m$.
Transducer ${\cal T}_m$ changes state if and only if one of the
following three mutually-exclusive conditions holds:
\begin{description}
\item{Type 1:} Transducer ${\cal T}_m$ is not in a $\Swait$ state,
  while all its successor transducers ${\cal T}_n$ are in $\Swait$
  states (where $m \prec n$).  ${\cal T}_m$ is free to make any transition.
\item{Type 2:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters a $\Smatch$ or $\Sinsert$ state, emitting a symbol and forcing ${\cal T}_m$ into a $\Smatch$ state so it can absorb that symbol.
\item{Type 3:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters the $\Send$ state, forcing ${\cal T}_m$ into the $\Send$ state as well.
\end{description}

A notational prescription for the allowed transitions may be found in \supptext.

\subsubsection*{How the ensemble generates multiple alignments}
